Quantum H theorem, variational entropy, and relaxation processes

### 2021-01-06 tex/quantum.tex
Rewording: start of sec. 3.2.1

In equilibrium (at $t\to\infty$), $dn_{\kappa}/dt=0$, hence from Eq.~(\ref{eq:changen})
%
\begin{equation}\label{eq:boseiferdirdistfnctsrc}
    \ln \frac{n_{\kappa}}{g_{\kappa}\pm n_{\kappa}}+\ln \frac{n_{\lambda}}{g_{\lambda}\pm n_{\lambda}}=\ln \frac{n_{\mu}}{g_{\mu}\pm n_{\nu}}+\ln \frac{n_{\nu}}{g_{\nu}\pm n_{\nu}}.
\end{equation}
%
Considering that the energy is not lost during the collsion, the Bose-Einstein or Fermi-Dirac distribution
functions can be recovered from Eq.~(\ref{eq:boseiferdirdistfnctsrc}):
%
\begin{equation}
   n_{\kappa}=\frac{g_{\kappa}}{\exp(\alpha+\beta\epsilon_{\kappa}\mp1)}.
\end{equation}
%

@@

Consider a diluted gas enclosed by a perfectly isolated vessel of volume $V$, with
total energy $E$, and total number of quantum particles $N$, which can be fermions or
bosons. For our purposes, the volume $V$ is divided into $K$ small cells, each of which
has a constant volume $\delta V_M=V/K$, a temperature $T_M$, a local energy $\epsilon_M$,
a local number of particles $\mathcal{N}_M$, and a local set of distribution functions,
$\{f_{Mn}(t)\}$. Each $f_{Mn}$ $\big(\equiv f(\vec r_M,\epsilon_{n},t)\big)$ represent the number
of particles contained in the $M$-th cell that occupy the energy level $\epsilon_n$.
We propose the following functional as an alternative $H$-functional
for quantum non-homogeneous diluted gases:
%
\begin{eqnarray}
    \mathcal{H} (t)&=&\sum_{M=1}^{K} \sum_{n} [ f_{Mn}(\epsilon_{n},t) \ln f_{Mn}(\epsilon_{n},t)\nonumber \\
    &&\pm (1 \mp f_{Mn}(\epsilon_{n},t)) \ln (1 \mp f_{Mn}(\epsilon_{n},t)) ]   \delta V_M \delta \epsilon_n \label{entropy}.
\end{eqnarray}
%
Here, the upper and lower sign refer to bosons and fermions respectively.
\color{blue}\commr{Also, the energy spectrum will be divided into $\delta \epsilon_n$ volumes}. We will use
the uppercase Latin index to denote cell numbers, and unless stated otherwise,
$M = 1, 2 . . . , K$. Without loss of generality, we choose these cells to have
local identical volumes (\textit{i.e.} $\delta V_M = V/K, \forall M$). Each
cell has its local chemical potential, $\mu_M$, local
variational functional, $\mathcal{H}_M$, local number of particles, $\mathcal{N}_M$, and
local energy, $\mathcal{E}_M$. In terms of the local distribution
function and local energy spectra, $\{\epsilon_{n}\}$, the last three
properties are given by:
%
\begin{eqnarray}
    \mathcal{H}_M &=&  \sum_{n} [ f_{Mn}(\epsilon_{n},t) \ln f_{Mn}(\epsilon_{n},t)\nonumber \\
    &&\pm (1 \mp f_{Mn}(\epsilon_{n},t)) \ln (1 \mp f_{Mn}(\epsilon_{n},t)) ] \de V_M \delta \epsilon_n \label{entropycell},\\
    {\mathcal{N}}_M&=& \sum_{n}f_{Mn}(\epsilon_{n} ,t) \de V_M \delta \epsilon_n, \nonumber \\
{\mathcal{E}}_M&=& \sum_{n}f_{Mn}(\epsilon_{n},t)\epsilon_{n} \de V_M \delta \epsilon_n.
\end{eqnarray}
%
When the system is in equilibrium, the local number of particles and the local
energy don't depend on the cell number, this is
%
\begin{equation}
   {\mathcal{N}}_M=\mathcal{N}\equiv \bar{\mathcal{N}}; \ \ \ \  {\mathcal{E}}_M=\mathcal{E}\equiv \bar{\mathcal{E}}.
\end{equation}
%
The particles in the system are considered to be free, \textit{i.e.} the
particles do not interact with each other.
Consequently, we can assume that the available quantum states do not depend on
the cell features,
therefore the energy spectrum, denoted by $\epsilon_n$, is the same for all
cells, \textit{i.e.} $\epsilon_{Mn} = \epsilon_n, \forall M$. Besides, the
system is subjected to the microcanonical restrictions
%
\begin{eqnarray}
    &&\sum_{M=1}^{K} \left[ \sum_{n}f_{Mn}(\epsilon_{n} ,t)\de \epsilon_n\right] \de V_M=\sum_{M=1}^{K} {\mathcal{N}}_{M} \de V_M=N; \nonumber \\
    &&\sum_{M=1}^{K}\left[ \sum_{n}f_{Mn}(\epsilon_{n},t)\epsilon_{n}\delta \epsilon_n\right] \de V_M=\sum_{M=1}^{K} {\mathcal{E}}_M \delta V_M=E, \label{restriccions1}
\end{eqnarray}
%
where $N$ and $E$ are the total particle number and the total energy of the
system respectively.

Maximizing the variational entropy $\mathcal{H}$ using the Lagrange multipliers method,
we obtain
%
\begin{eqnarray}
&&\ln \left(\frac{1\mp f_{Mn}(\epsilon_{n},t)}{f_{Mn}(\epsilon_{n},t)} \right)=-\alpha_M(t)-\beta_M(t) \epsilon_{n}, \label{relation}\\ &&\Rightarrow \ \ f_{Mn}(\epsilon_{n},t)=\frac{1}{e^{-\alpha_M(t)-\beta_M(t) \epsilon_{n}}\pm 1} \equiv \bar{f}_{Mn}(\epsilon_{n},t) \label{distributionequilibrium}.
\end{eqnarray}
%
where $\alpha_M(t), \beta_M(t)$ are the Lagrange multipliers. We can obtain
from the variational procedure that
%
\begin{equation}
    \alpha_M\propto \frac{\partial \ln \Omega_M}{\partial \mathcal{N}_M}, \ \ \ \beta_M\propto \frac{\partial \ln \Omega_M}{\partial \mathcal{E}_M},\label{multipliers}
\end{equation}
%
where we used that the $H$-functional in equilibrium is proportional to the
entropy of the system.
In other hand, in the expression (\ref{multipliers}), we expect that the
variation of the logarithm with the number of particles by cells is indifferent
from a particular cell. With this, we conclude 
%
\begin{equation}
    \alpha_M=\alpha; \ \ \ \ \beta_M=\beta.
\end{equation}
%
In equilibrium, the energy and the Lagrange multiplier do not depend on the
cell number and time, therefore
%
\begin{equation}
    \bar f_{Mn}(\epsilon_{n},t)=\bar f_n(\epsilon_{n}) =\frac{1}{e^{-\alpha-\beta \epsilon_n}\pm 1}.
\end{equation}
%
If one compares those multipliers with the mean energy and the mean particle
number from the statistical mechanic's results, we obtain
%
\begin{equation}
    \alpha=\frac{\mu}{kT}\equiv \bar{\alpha}; \ \ \ \ \beta=-\frac{1}{kT}\equiv \bar{\beta},
\end{equation}
%
and finally
%
\begin{eqnarray}
    \bar{f}_{n}(\epsilon_{n})&=&\frac{1}{e^{(\frac{{\epsilon_n}-\bar{\mu}}{kT})}\pm 1}\equiv \bar{f}_{n}.
\end{eqnarray}
%
We proved, when the system is in equilibrium, the Lagrange multipliers don't
depend on the cell number. In other words, Lagrange multipliers are homogeneous
in the system.

>>>>

In equilibrium (at $t\to\infty$), $dn_{\kappa}/dt=0$, hence from Eq.~(\ref{eq:changen})
%
\begin{equation}\label{eq:boseiferdirdistfnctsrc}
    \ln \frac{n_{\kappa}}{g_{\kappa}\pm n_{\kappa}}+\ln \frac{n_{\lambda}}{g_{\lambda}\pm n_{\lambda}}=\ln \frac{n_{\mu}}{g_{\mu}\pm n_{\nu}}+\ln \frac{n_{\nu}}{g_{\nu}\pm n_{\nu}}.
\end{equation}
%
Considering that the energy is conserved during the collsion, the Bose-Einstein or Fermi-Dirac distribution
functions can be recovered from Eq.~(\ref{eq:boseiferdirdistfnctsrc}):
%
\begin{equation}
   n_{\kappa}=\frac{g_{\kappa}}{\exp(\alpha+\beta\epsilon_{\kappa}\mp1)}.
\end{equation}
%

@@

Consider a diluted gas enclosed by a perfectly isolated vessel of volume $V$, with
total energy $E$, and total number of quantum particles $N$, which can be free fermions or
bosons. For our purposes, the volume $V$ is divided into $K$ small cells, each of which
has a constant volume $\delta V_M=V/K$ ($M=1,\,\dots\,,K$), a
local temperature $T_M$, a local energy $\epsilon_M$,
a local number of particles $\mathcal{N}_M$, and a local set of distribution functions,
$\{f_{Mn}(t)\}$. Hereafter we will use the short-hand notation:
%
\begin{equation}
   f_{Mn}(t)\equiv f(\vec r_M,\epsilon_{n},t),
\end{equation}
%
where $\vec r_M$ is the radius vector pointing at the center of the $M$-th cell.
$f_{Mn}(t)$ represent the number
of particles contained in the $M$-th cell that occupy the energy level $\epsilon_n$ at
time $t$. Since the particles are considered to be free, the energy levels should not
depend on the cell properties, \textit{i.e.} the energy spectrum, $\{\epsilon_n\}$, is the same for all
cells; thus there is no need to label $\epsilon_n$ with an index $M$.

We propose the following functional as an alternative $H$-functional
for quantum non-homogeneous diluted gases:
%
\begin{eqnarray}
    \mathcal{H} (t)&=&\sum_{M=1}^{K} \sum_{n} \bigg[ f_{Mn}(t) \ln f_{Mn}(t)\bigg.\nonumber \\
    &&\bigg.\pm \Big(1 \mp f_{Mn}(t)) \ln (1 \mp f_{Mn}(t)\Big) \Big]   \delta V_M\label{entropy}.
\end{eqnarray}
%
Here, the upper and lower sign refer to bosons and fermions respectively.
In addition, when needed, each cell will have an associated local chemical potential,
$\alpha_M$, and its local $H$-functional, which is defined by:
%
\begin{equation}\label{eq:qHMdef}
   \mathcal{H}_M(t)=\sum_{n} \bigg[ f_{Mn}(t) \ln f_{Mn}(t)
   \pm \Big(1 \mp f_{Mn}(t)\Big) \ln \Big(1 \mp f_{Mn}(t)\Big) \bigg] \de V_M.
\end{equation}
%
Therefore, $\mathcal{N}_M$ and $\mathcal{E}_M$ as functions of time are given by:
%
\begin{subequations}
\begin{equation}
    {\mathcal{N}}_M(t) = \sum_{n}f_{Mn}(t) \delta V_M
\end{equation}
%
and
%
\begin{equation}
\mathcal{E}_M(t) = \sum_{n}f_{Mn}(t)\epsilon_{n} \de V_M
\end{equation}
\end{subequations}
%
which are, in addtion, constrained by the global microcanonical restrictions
%
\begin{subequations}
\begin{equation}\label{eq:qGlobRestN}
    \sum_{M=1}^{K}\left[\sum_{n}f_{Mn}(t)\right]\delta  V_M
    =\sum_{M=1}^{K}\mathcal{N}_{M}(t)\delta V_M=N
\end{equation}
%
and
%
\begin{equation}\label{eq:qGlobRestE}
    \sum_{M=1}^{K}\left[\sum_{n}f_{Mn}(t)\epsilon_{n}\right]\delta V_M
    =\sum_{M=1}^{K}\mathcal{E}_M(t)\delta V_M=E. 
\end{equation}
\end{subequations}
%

Applying the variational method to $\mathcal{H}$, and using the using the
Lagrange multipliers
$\{\alpha_M\}$ and $\{\beta_M\}$ it is straightforward to obtain
(see also text around Eq.~(\ref{eq:deltaHpdeltafpj})):
%
\begin{equation}\label{eq:relation}
\ln \left(\frac{1\mp f_{Mn}(t)}{f_{Mn}(t)} \right)=-\alpha_M(t)-\beta_M(t) \epsilon_{n},
\end{equation}
%
and solving for $f_{Mn}(t)$ yields
%
\begin{equation}\label{eq:qfMn}
f_{Mn}(t)=\frac{1}{\exp\big(-\alpha_M(t)-\beta_M(t) \epsilon_{n}\big)\pm 1}.
\end{equation}

%============================================================
\subsubsection{Properties of $\mathcal{H}$ in equilibrium}
%============================================================
If the system is in equilibrium (at $t\to\infty$), the local number of particles and the local
energy do not depend on the cell number and they should be homogeneous, this is \commr{[This needs a rationale\dots]}
%
\begin{subequations}
\begin{equation}
   {\mathcal{N}}_M(t\to\infty)\equiv \bar{\mathcal{N}}=N/K
\end{equation}
%
and
%
\begin{equation}
	{\mathcal{E}}_M(t\to\infty)\equiv \bar{\mathcal{E}}=E/K.
\end{equation}
%
\end{subequations}
%

\color{blue}

%
\commr{We can obtain
from the variational procedure that
%
\begin{equation}
    \alpha_M\propto \frac{\partial \ln \Omega_M}{\partial \mathcal{N}_M}, \ \ \ \beta_M\propto \frac{\partial \ln \Omega_M}{\partial \mathcal{E}_M},\label{multipliers}
\end{equation}
%
[What is $\Omega_M$?]
}

where we used that the $H$-functional in equilibrium is proportional to the
entropy of the system.
In other hand, in the expression (\ref{multipliers}), we expect that the
variation of the logarithm with the number of particles by cells is indifferent
from a particular cell. With this, we conclude 
%
\begin{equation}
    \alpha_M=\alpha; \ \ \ \ \beta_M=\beta.
\end{equation}
%
In equilibrium, the energy and the Lagrange multiplier do not depend on the
cell number and time, therefore
%
\begin{equation}
    \bar f_{Mn}(\epsilon_{n},t)=\bar f_n(\epsilon_{n}) =\frac{1}{e^{-\alpha-\beta \epsilon_n}\pm 1}.
\end{equation}
%
If one compares those multipliers with the mean energy and the mean particle
number from the statistical mechanic's results, we obtain
%
\begin{equation}
    \alpha=\frac{\mu}{kT}\equiv \bar{\alpha}; \ \ \ \ \beta=-\frac{1}{kT}\equiv \bar{\beta},
\end{equation}
%
and finally
%
\begin{eqnarray}
    \bar{f}_{n}(\epsilon_{n})&=&\frac{1}{e^{(\frac{{\epsilon_n}-\bar{\mu}}{kT})}\pm 1}\equiv \bar{f}_{n}.
\end{eqnarray}
%
We proved, when the system is in equilibrium, the Lagrange multipliers don't
depend on the cell number. In other words, Lagrange multipliers are homogeneous
in the system.
